---
title: "Propensity Score Matching in R (My first post in blogger)"
author: "Andrés Gutiérrez"
date: "2014-11-14"
categories:
  - "Policy Evaluation"
  - "R"
---

My first post in blogger is about Propensity Score Matching (PSM) in R. In this post I will introduce the R packages MatchIt and Zelig. The first one is devoted to perform different PSM algorithms and the other one is used in order to estimate the average treatment effect (ATE) and the effect of the treatment on the treated (ATT). First of all, I will simulate a population where some individuals receive the treatment, and the remaining ones do not.

```{r}
#| error: false
rm(list=ls())
```
```{r}
#| error: false
library(MatchIt)
```
```{r}
#| error: false
# Zelig is no longer available on CRAN; the equivalent analyses below use lm().
```
```{r}
#| error: false
N <- 1000
```
```{r}
#| error: false
x1 <- round(rnorm(N,20,4))
```
```{r}
#| error: false
x2 <- round(rnorm(N,80,10))
```
```{r}
#| error: false
f <- 1 + 1*x1 - 0.28*x2
```
```{r}
#| error: false
p <- exp(f)/(1+exp(f))
plot(p)
```
```{r}
#| error: false
z <- rbinom(N,1, p)
table(z)
```
```{r}
#| error: false
y <- rep(0, times=N)
```
```{r}
#| error: false
# Different values for y over z# The mean difference for y is 5000
```
```{r}
#| error: false
y[z == 1] <- 9000 + 2*x1[z == 1] + 0.5*x2[z == 1]
```
```{r}
#| error: false
y[z == 0] <- 4000 + 2*x1[z == 0] + 0.5*x2[z == 0]
```
```{r}
#| error: false
y <- jitter(y, amount=1000,factor=2)
```
```{r}
#| error: false
Data <- data.frame(y=y, x1=x1, x2=x2, z)
```
```{r}
#| error: false
# The treatment population
```
```{r}
#| error: false
Treats <- subset(Data, z == 1)
```
```{r}
#| error: false
colMeans(Treats)
```
```{r}
#| error: false
# The control population
```
```{r}
#| error: false
Control <- subset(Data, z == 0)
colMeans(Control)
```
```{r}
#| error: false
# Computing the real effect of the treatment
```
```{r}
#| error: false
model <- lm(y ~ z + x1 + x2,data=Data)
```
```{r}
#| error: false
Effect <- model$coeff[2]
Effect
```
Then I will run a logistic regression in order to estimate the propensity scores. That is the probability of receiving the treatment.

```{r}
#| error: false
# The propensity scores
```
```{r}
#| error: false
ps.model <- glm(z ~ x1 + x2,family = binomial(link = "logit"),data = Data)
```
```{r}
#| error: false
plot(ps.model$fitted)
summary(ps.model)
```
```{r}
#| error: false
Data$ps <- ps.model$fitted
table(Data$z)
```
```{r}
#| error: false
hist(Data$ps[z==1])
```
```{r}
#| error: false
hist(Data$ps[z==0])
```
After that, one may use different matching algorithms. Exact matching is a technique used to match individuals based on the exact values of covariates.

```{r}
#| error: false
# Exact matching
```
```{r}
#| error: false
m1.ps <- matchit(z ~ x1 + x2, method = "exact", data = Data)
```
```{r}
#| error: false
summary(m1.ps, covariates = T)
```
```{r}
#| error: false
m1.data <- match.data(m1.ps)
```
Nearest neighbour matching (NNM) is an algorithm that matches individuals with controls (it could be more two or more controls per treated unit) one by one based on a distance.

```
#Nearest neighbor matching
```

```{r}
#| error: false
m3.ps <- matchit(z ~ x1 + x2, method = "nearest", ratio = 1, data = Data)
```
```{r}
#| error: false
summary(m3.ps, covariates = T)
```
```{r}
#| error: false
plot(m3.ps)
```
```{r}
#| error: false
plot(m3.ps, type="jitter")
```
```{r}
#| error: false
plot(m3.ps, type="hist")
```
```{r}
#| error: false
m3.data <- match.data(m3.ps)
```
```{r}
#| error: false
m3.ps$match.matrix
```
```{r}
#| error: false
Pairs <- cbind(Data[row.names(m3.ps$match.matrix),], Data[m3.ps$match.matrix,])
```
```{r}
#| error: false
Pairs
```
Optimal matching solves the problem inherent to the NNM, that matches units one at a time, without taking into account any global (over the entire population) distance measure. This way, with NNM the last treated unit could could match an inappropriate unit.

```{r}
#| error: false
#Optimal matching
```
```{r}
#| error: false
m5.ps <- matchit(z ~ x1 + x2, method = "optimal", data = Data)
```
```{r}
#| error: false
summary(m5.ps, covariates = T)
```
```{r}
#| error: false
plot(m5.ps)
```
```{r}
#| error: false
plot(m5.ps, type="jitter")
```
```{r}
#| error: false
plot(m5.ps, type="hist")
```
```{r}
#| error: false
m5.data <- match.data(m5.ps)
```
```{r}
#| error: false
# m5.data[(m5.data$subclass == 1),]
```
Full matching forms subgroups in an optimal way.

```{r}
#| error: false
#Full matching
```
```{r}
#| error: false
m4.ps <- matchit(z ~ x1 + x2, method = "full", data = Data)
```
```{r}
#| error: false
summary(m4.ps, covariates = T)
```
```{r}
#| error: false
plot(m4.ps)
```
```{r}
#| error: false
plot(m4.ps, type="jitter")
```
```{r}
#| error: false
plot(m4.ps, type="hist")
```
```{r}
#| error: false
m4.data <- match.data(m4.ps)
```
Caliper matching is a refinement of the NNM. It defines a constraint over the individual distant measure. So, it minimises the probability of bad matching.

```{r}
#| error: false
# Caliper matching
```
```{r}
#| error: false
sd.ps <- sd(Data$ps)
m8.ps <- matchit(z ~ x1 + x2, method = "nearest", caliper = 0.25 * sd.ps, data = Data)
```
```{r}
#| error: false
summary(m8.ps, covariates = T)
```
```{r}
#| error: false
plot(m8.ps)
```
```{r}
#| error: false
plot(m8.ps, type="jitter")
```
```{r}
#| error: false
plot(m8.ps, type="hist")
```
```{r}
#| error: false
m8.data <- match.data(m8.ps)
```
```{r}
#| error: false
m8.data.control <- match.data(m8.ps,"control")
```
```{r}
#| error: false
m8.data.treat <- match.data(m8.ps,"treat")
```
```{r}
#| error: false
m8.ps$match.matrix
```
```{r}
#| error: false
Pairs <- cbind(Data[row.names(m8.ps$match.matrix),], Data[m8.ps$match.matrix,])
```
```{r}
#| error: false
Pairs
```
Finally, with the matched data, we estimate the average treatment effects, via automatic simulation, with the Zelig package.

```{r}
#| error: false
# Treatment Effects
```
```{r}
#| error: false
z1.ps <- lm(y ~ z + x1 + x2, data = m8.data)
```
```{r}
#| error: false
summary(z1.ps)
```
```{r}
#| error: false
x <- transform(m8.data, z = 0)
x1 <- transform(m8.data, z = 1)
```
```{r}
#| error: false
s.out <- mean(predict(z1.ps, newdata = x1) - predict(z1.ps, newdata = x))
```
```{r}
#| error: false
s.out
```
```{r}
#| error: false
barplot(s.out, main = "Estimated average treatment effect")
```
Zelig package also allows to estimate the effect on the treated, and the effect on the controls.

```{r}
#| error: false
# Average Treatment Effect on the Controls
```
```{r}
#| error: false
z1c.ps <- lm(y ~ x1 + x2, data = m8.data.control)
```
```{r}
#| error: false
summary(z1c.ps)
```
```{r}
#| error: false
x.out1 <- predict(z1c.ps, newdata = m8.data.treat)
```
```{r}
#| error: false
s.out1 <- mean(m8.data.treat$y - x.out1)
s.out1
```
```{r}
#| error: false
barplot(s.out1, main = "Estimated effect on the treated")
```
```{r}
#| error: false
# Average Treatment Effect on the Treated
```
```{r}
#| error: false
z1t.ps <- lm(y ~ x1 + x2, data = m8.data.treat)
```
```{r}
#| error: false
summary(z1t.ps)
```
```{r}
#| error: false
x.out2 <- predict(z1t.ps, newdata = m8.data.control)
```
```{r}
#| error: false
s.out2 <- mean(x.out2 - m8.data.control$y)
```
```{r}
#| error: false
s.out2
```
```{r}
#| error: false
barplot(s.out2, main = "Estimated effect on the controls")
```
I hope that you found this post valuable. Feel free to reproduce the code wherever you want.
