---
title: "Automatic output format in Rmarkdown"
author: "Andrés Gutiérrez"
date: "2017-11-19"
categories:
  - "R"
  - "Software"
---

I am writing a Rmarkdown document with plenty of tables, and I want them in a decent format, e.g. kable. However I don't want to format them one by one. For example, I have created the following data frame in dplyr

```{r}
#| error: false
library(dplyr)
data2 <- data.frame(uf = sample(state.abb[1:8], 100, replace = TRUE))
data2 %>% group_by(uf) %>%
  summarise(n = n(), ) %>%
  arrange(desc(n))
```
One solution to the output format of this data frame would be to name it as an object in R, and then give it a format by using the kable function.

```{r}
#| error: false
t1 <- data2 %>%
  group_by(uf) %>%
  summarise(n = n(), ) %>% arrange(desc(n))

knitr::kable(t1)
```
However, if your document has hundreds of these queries and you need a faster way to compile the document, while keeping the kable style automatically, avoiding giving a name to the data frame and even avoiding to call the kable function over that name, you can use the [printr package](https://yihui.name/printr/). Just add the following piece of code inside a chunk at the beginning of your document and *voilá.*

```{r}
#| error: false
library(printr)
```
Now, all of your data frames will have a decent style, and you do not need to worry about this issue. For example, I have knitted a presentation by using *printr*and the first code in this post, and this is the result:
